Parsing Postfix log files

### 2009-03-09 doc/logparser-architecture.tex
Changes following more comments from Carl

\subsection{Attaching Conditions To Rules}

@@

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua.  If an implementation is going
to use conditions so complex that they will require a Turing-equivilent
programming language to implement them, the design may need to be
re-thought, including the decision to use this architecture --- there may
be other architectures more suitable.

@@

An example of how rule conditions can be useful is parsing C-style
comments, which start with \texttt{/*} and end with \texttt{*/}; the start
and end tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These rules are somewhat simplified, e.g.\ rules one and two will recognise
the comment start token even if it is within a quoted string, which is
incorrect.  Rules one and two will be used when the parser is parsing code,
not comments: rule one recognises a comment that is contained within one
line and leaves the parser's state unchanged; rule two recognises the start
of a comment and changes the parser's state to parsing comments.  Rules
three and four will be used when the parser is parsing a comment: rule
three recognises the end of a comment and switches the parser's state back
to parsing code; rule four recognises a comment line without an end token
and keeps the parser's state unchanged.  It is important that the rules are
applied in the order listed in \tableref{Rules to parse C-style comments};
\sectionref{rules in architecture} has explained how this is achieved, and
\sectionref{overlapping rules in architecture} will discuss overlapping
rules.

@@

When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Depending on the conditions employed a rule
may overlap with multiple rules, and rule overlap is not transitive, e.g.\
given these three conditions:

@@

    \item $total < 10$

    \item $total > 20$

    \item $total < 30$

@@

For this approach to be more efficient than the naive approach $y$ must be
significantly lower than $x$.  If $y$ is higher than $x$ then the opposite
approach should be taken: check for regex overlap first, then check
for condition overlap only between pairs of rules with overlapping regexes.

@@

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
in-depth discussion of nested quantifiers, backtracking, alteration, and
capturing groups.  Pathological regexes which consume excessive CPU time
can be difficult to detect, both by visual inspection and by machine
inspection, but if a regex is converted to a \acronym{FA} or the internal
representation used by the regex engine, it may be possible to determine if
there are nested quantifiers or other troublesome constructs present.
Modern regex engines have addressed many of these problems, e.g.\ the regex
to match double quoted strings given above fails immediately with Perl
5.10, regardless of the input length, because the regex engine looks for
both of the required double quotes first.  Similarly Perl 5.10's regex
engine optimises alterations starting with literal text into a trie, which
has matching time proportional to the length of the alternatives, rather
than the number of alternatives.  Perl regexes can use \verb!(?>pattern)!,
which matches \verb!pattern! the first time the regex engine passes over
it, and does not change what it originally matched if the regex engine
backtracks over it, so troublesome regexes can use this to reduce the
impact of backtracking; Prolog users will notice a similarity to the
\verb'!' (cut) operator.  A presentation showing some of Perl 5.10's new
regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors too.
The three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework, unsurprisingly, was quite quick to document, because it is the
simplest of the three components.  The actions are also quite easy to
understand, because the architecture does not impose any structure or
requirements upon them: parser authors are free to do anything they want
within an action.  The architecture's support for cascaded parsing was
described in the actions section, with an example to illustrate how it can
be useful.  The rules section was the longest, because although the rules
appear conceptually simple --- recognise an input and specify the action
that handles that input --- they have subtle behaviour that needs to be
clearly explained.  When extending a ruleset a decision needs to be taken
about whether the input should be recognised by extending an existing rule,
by adding a new rule to an existing rule category, or by adding a new rule
category, rule, and action.  Rules can have conditions attached to them,
restricting the set of rules used to recognise an input; the complexity of
the conditions used greatly influences the difficulty of writing a correct
ruleset or understanding and extending an existing ruleset.  Overlapping
rules are frequently a requirement in a parser, and their use can greatly
simplify some rules, but they can be a source of bugs because they can
recognise inputs unexpectedly.  The framework does not try to detect
overlapping rules, because overlap amongst rules may be valid and is quite
often intentional; that responsibility falls to the author of the rules.
The difficulty involved in detecting overlap is proportional to the
complexity of a ruleset's regexes and conditions, and may be possible for a
human yet intractable or impossible for a program.  The rules section
finishes with a discussion of pathological rules, concentrating on
pathological regexes.

>>>>

\newpage{} % XXX
\subsection{Attaching Conditions To Rules}

@@

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua.  If an implementation is going
to use conditions so complex that they will require a Turing-equivalent
programming language to implement them, the design may need to be
re-thought, including the decision to use this architecture --- there may
be other architectures more suitable.

@@

An example of how rule conditions can be useful is parsing C-style
comments, which start with \texttt{/*} and end with \texttt{*/}; the start
and end tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These rules are somewhat simplified, e.g.\ rules one and two will recognise
the comment start token even if it is within a quoted string, which is
incorrect; the rules could be extended with new rules to parse quoted
strings, excluding comments from within strings.

Rules one and two will be used when the parser is parsing code, not
comments: rule one recognises a comment that is contained within one line
and leaves the parser's state unchanged; rule two recognises the start of a
comment and changes the parser's state to parsing comments.  Rules three
and four will be used when the parser is parsing a comment: rule three
recognises the end of a comment and switches the parser's state back to
parsing code; rule four recognises a comment line without an end token and
keeps the parser's state unchanged.  It is important that the rules are
applied in the order listed in \tableref{Rules to parse C-style comments}
because rule 2 overlaps with rule 1, so rule 1 must be used first to
correctly parse comments that are contained within one line;
\sectionref{rules in architecture} has explained how this is achieved.
\sectionref{overlapping rules in architecture} will discuss the benefits
and difficulties of using overlapping rules.

@@

When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Depending on the conditions employed a rule
may overlap with multiple rules, and rule overlap is not transitive, e.g.\
given these three artificial conditions:

@@

    \item total $<$ 10

    \item total $>$ 20

    \item total $<$ 30

@@

In the worst case no overlap will be found between conditions, so all pairs
or regexes must be checked for overlap and $|o|$ above will be equal to
$n^{2}$.  For this approach to be more efficient than the naive approach
$y$ must be significantly lower than $x$.  If $y$ is higher than $x$ then
the checks for overlap should be performed in the opposite order: check for
regex overlap first, then check for condition overlap only between pairs of
rules with overlapping regexes.

@@

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity for a string of length $n$ is $O(2^{n})$;
see~\cite{mastering-regular-expressions} for in-depth discussion of nested
quantifiers, backtracking, alteration, and capturing groups.  Pathological
regexes which consume excessive CPU time can be difficult to detect, both
by visual inspection and by machine inspection, but if a regex is converted
to a \acronym{FA} or the internal representation used by the regex engine,
it may be possible to determine if there are nested quantifiers or other
troublesome constructs present.  Modern regex engines have addressed many
of these problems, e.g.\ the regex to match double quoted strings given
above fails immediately with Perl 5.10, regardless of the input length,
because the regex engine looks for both of the required double quotes
first.  Similarly Perl 5.10's regex engine optimises alterations starting
with literal text into a trie, which has matching time proportional to the
length of the alternatives, rather than the number of alternatives.  Perl
regexes can use \verb!(?>pattern)!, which matches \verb!pattern! the first
time the regex engine passes over it, and does not change what it
originally matched if the regex engine backtracks over it, so troublesome
regexes can use this to reduce the impact of backtracking; Prolog users
will notice a similarity to the \verb'!' (cut) operator.  A presentation
showing some of Perl 5.10's new regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors too.
The three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework provides a number of support functions and manages the parsing
process, enabling simple rules and actions to be written.  The actions are
quite easy to understand, because the architecture does not impose any
structure or requirements upon them: parser authors are free to do anything
they want within an action.  The architecture's support for cascaded
parsing was described in the actions section, with an example to illustrate
how it can be useful for general parsing.  The rules section was the
longest, because although the rules appear conceptually simple ---
recognise an input and specify the action that handles that input --- they
have subtle behaviour that needs to be clearly explained.  When extending a
ruleset a decision needs to be taken about whether the input should be
recognised by extending an existing rule, by adding a new rule to an
existing rule category, or by adding a new rule category, rule, and action.
Rules can have conditions attached to them, restricting the set of rules
used to recognise an input; the complexity of the conditions used greatly
influences the difficulty of writing a correct ruleset or understanding and
extending an existing ruleset.  Overlapping rules are frequently a
requirement in a parser, and their use can greatly simplify some rules, but
they can be a source of bugs because they can recognise inputs
unexpectedly.  The framework does not try to detect overlapping rules,
because overlap amongst rules may be valid and is quite often intentional;
that responsibility falls to the author of the rules.  The difficulty
involved in detecting overlap is proportional to the complexity of a
ruleset's regexes and conditions, and may be possible for a human yet
intractable or impossible for a program.  The rules section finishes with a
discussion of pathological rules, concentrating on pathological regexes.

### 2009-03-09 doc/logparser-introduction.tex
Changes following more comments from Carl

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview of the architecture,
then describing the components of the architecture in detail.

>>>>

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview of the architecture,
then describing each of the components of the architecture in detail.
